---
title: "CCREPE: Compositionality Corrected by PErmutation and REnormalization"
author: "Emma Schwager, George Weingart, Craig Bielski, Curtis Huttenhower"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{ccrepe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`ccrepe` is a package for analysis of sparse compositional data. Specifically, it determines the significance of association between features in a composition, using any similarity measure (e.g. Pearson correlation, Spearman correlation, etc.) The CCREPE methodology stands for Compositionality Corrected by Renormalization and Permutation. The package also provides a novel similarity measure, the N-dimensional checkerboard score (NC-score), particularly appropriate to compositions binned from continuous data.

# ccrepe

`ccrepe` is the main package function. It calculates compositionality-corrected p-values and q-values for a user-selected similarity measure, operating on either one or two input matrices.

## Example 1

An example of how to use `ccrepe` with one dataset.

```{r example1}
library(ccrepe)
data <- matrix(rlnorm(40,meanlog=0,sdlog=1),nrow=10,ncol=4)
data[,1] = 2*data[,2] + rnorm(10,0,0.01)
data.rowsum <- apply(data,1,sum)
data.norm <- data/data.rowsum

test.input <- data.norm
dimnames(test.input) <- list(c(
 "Sample 1", "Sample 2","Sample 3","Sample 4","Sample 5",
 "Sample 6","Sample 7","Sample 8","Sample 9","Sample 10"),
 c("Feature 1", "Feature 2", "Feature 3","Feature 4"))

test.output <- ccrepe(x=test.input, iterations=20, min.subj=10)

par(mfrow=c(1,2))
plot(data[,1],data[,2],xlab="Feature 1",ylab="Feature 2",main="Non-normalized")
plot(data.norm[,1],data.norm[,2],xlab="Feature 1",ylab="Feature 2",
     main="Normalized")
test.output
```

## Example 2

An example of how to use `ccrepe` with two datasets.

```{r example2}
data2 <- matrix(rlnorm(105,meanlog=0,sdlog=1),nrow=15,ncol=7)
aligned.rows <- c(seq(1,4),seq(6,9),11,12)
data2[aligned.rows,1] <-  2*data[,3] + rnorm(10,0,0.01)
data2.rowsum <- apply(data2,1,sum)
data2.norm <- data2/data2.rowsum
test.input.2 <- data2.norm

dimnames(test.input) <- list(paste("Sample",seq(1,10)),paste("Feature",seq(1,4)))
dimnames(test.input.2) <- list(paste("Sample",c(seq(1,4),11,seq(5,8),12,9,10,13,14,15)),paste("Feature",seq(1,7)))

test.output.two.datasets <- ccrepe(x=test.input, y=test.input.2, iterations=20, min.subj=10)

par(mfrow=c(1,2))
plot(data2[aligned.rows,1],data[,3],xlab="dataset 2: Feature 1",ylab="dataset 1: Feature 3",main="Non-normalized")
plot(data2.norm[aligned.rows,1],data.norm[,3],xlab="dataset 2: Feature 1",ylab="dataset 1: Feature 3",
     main="Normalized")
test.output.two.datasets
```

## Example 3

An example of how to use `ccrepe` with `nc.score` as the similarity score.

```{r example3}
test.output.nc.score <- ccrepe(x=test.input, sim.score=nc.score, iterations=20, min.subj=10)

par(mfrow=c(1,2))
plot(data[,1],data[,2],xlab="Feature 1",ylab="Feature 2",main="Non-normalized")
plot(data.norm[,1],data.norm[,2],xlab="Feature 1",ylab="Feature 2",
     main="Normalized")
test.output.nc.score
```

# nc.score

The `nc.score` similarity measure is an N-dimensional extension of the checkerboard score suited to ecological relative abundance measurements.

## Example 1

An example of using `nc.score` to get a single similarity score or a matrix.

```{r nc_example1}
test.output.matrix <- nc.score(x=test.input)
test.output.num    <- nc.score(x=test.input[,1],y=test.input[,2])

par(mfrow=c(1, 2))
plot(data[,1],data[,2],xlab="Feature 1",ylab="Feature 2",main="Non-normalized")
plot(data.norm[,1],data.norm[,2],xlab="Feature 1",ylab="Feature 2",
     main="Normalized")
test.output.matrix
test.output.num
```
